 

### 2020-09-22 vertical-thermal-review.Rmd
added Lawren Sack to the author list

**Authors (so far, not necessarily in final order):** 
Nidhi Vinod^1^,
Martijn Slot^2^,
Ian McGregor^3^,
Elsa Ordway,
Marielle N. Smith^3^,
Tyeen Taylor,
Kristina J. Anderson-Teixeira^1,2^*

@@

Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
There is also increasing severity and frequency of heat waves (@perkinsIncreasingFrequencyIntensity2012).
This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
These changes, alongside climate change-driven disturbances such as drought, deforestation and fire, may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008, @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015].

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Larger trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate. (Nakamura et al. 2017, Ozanne et al. 2003)  
With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (McDowell et al.2020).

**The structure and microclimates of global forests are changing due to unprecedented disturbance rates.**
Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Additionally, loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components (Zellweger et al. 2020; Stark et al. 2020 ), this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]

>>>>

**Authors (so far, not necessarily in final order):** 
Nidhi Vinod^1^,
Martijn Slot^2^,
Ian McGregor^3^,
Elsa Ordway,
Marielle N. Smith^3^,
Tyeen Taylor,
Lawren Sack,
Kristina J. Anderson-Teixeira^1,2^*

@@

Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
There is also increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012].
This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
These changes, alongside climate change-driven disturbances such as drought, deforestation and fire, may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008, @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015].

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Taller trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate. (Nakamura et al. 2017, Ozanne et al. 2003)  
With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (McDowell et al.2020).

**The structure and microclimates of global forests are changing due to unprecedented disturbance rates.**
Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components (Zellweger et al. 2020; Stark et al. 2020 ), this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]
